---
title: "Annuals and perennials across Australia's rainfall gradient — results"
format: html
execute:
  echo: false
  warning: false
---

```{r}
#| label: setup
#| include: false
library(targets)
library(dplyr)
library(knitr)
```

This document reports the results. It performs no computation: every figure,
table and model below is read from the `targets` pipeline with `tar_read()`.
Run `targets::tar_make()` first if anything is out of date.

Reading rather than recomputing is deliberate. When this document did its own
work the same analysis existed both here and in `R/`, and the two copies could
drift apart. There is now one definition of each result, and this file displays
it.

```{r}
#| label: check
if (length(tar_outdated()) > 0) {
  warning("Some targets are out of date. Run targets::tar_make() before trusting this report.")
}
```

# Data summary

```{r}
#| label: tbl-counts
#| tbl-cap: "Species counts by origin and life history."
tar_read(table1) |> kable()
```

# Figures

## Figure 1. Climate space for Australia

Mean annual precipitation and precipitation seasonality across the continent,
the climate space they define, and total species richness.

```{r}
#| label: fig-1
include_graphics(tar_read(fig1_png))
```

## Figure 2. Richness of annuals and perennials, native and introduced

```{r}
#| label: fig-2
include_graphics(tar_read(fig2_png))
```

## Figure 3. Richness and the annual fraction against climate

Fitted at the primary effort cutoff of 500 records per cell.

```{r}
#| label: fig-3
include_graphics(tar_read(fig3_png))
```

```{r}
#| label: tbl-fig3-stats
#| tbl-cap: "Coefficients for the three models in Figure 3."
tar_read(fig3_stats_table_500) |> kable()
```

### Residual standard error by group

Residual standard error is used in preference to R², because R² is sensitive to
the total variance in the response — groups with flatter responses yield a lower
R² even where the model fits equally tightly in absolute terms. RSE gives a
directly comparable measure of typical absolute deviation between observed and
fitted log10 richness. It requires a separate model per group.

```{r}
#| label: tbl-rse
#| tbl-cap: "Residual standard error, one model per origin and life-history group."
tar_read(rse_fits) |> kable()
```

## Figure 4. Annual fraction across geography and climate space

```{r}
#| label: fig-4
include_graphics(tar_read(fig4_png))
```

## Figure 5. Introduced fraction of annuals, and population accessibility

```{r}
#| label: fig-5
include_graphics(tar_read(fig5_png))
```

```{r}
#| label: tbl-prop-invasive
#| tbl-cap: "Model fit for effect of population accessibility on likelihood an annual is introduced (versus native). Supplementary Table S2 in the manuscript."
tar_read(tableS2) |>
  dplyr::select(parameter, estimate_se, z_statistic, p_value) |>
  kable(col.names = c("Parameter", "estimate ± SE", "z-statistic", "p-value"))
```

# Supplementary figures

## Figure S1. Australian against global climate space

```{r}
#| label: fig-s1
include_graphics(tar_read(figS1_png))
```

## Figures S2–S4. Sensitivity to sampling effort per grid cell

Figure 3 re-run at four effort cutoffs, to check that the patterns do not depend
on the inclusion of sparsely sampled cells.

```{r}
#| label: fig-cutoffs
include_graphics(c(tar_read(figS_png_100), tar_read(figS_png_250),
                   tar_read(figS_png_500), tar_read(figS_png_1000)))
```

## Figure S5. Sampling effort in climate space and geographically

```{r}
#| label: fig-s5
include_graphics(tar_read(figS5_png))
```

## Figure S6. Effort diagnostics

```{r}
#| label: fig-s6
include_graphics(tar_read(figS6_png))
```

```{r}
#| label: tbl-effort
#| tbl-cap: "Model of sampling effort against climate."
parameters::parameters(tar_read(effort_model)) |> as.data.frame() |> kable()
```

# Extended data

## Figure E2. Effort-standardised richness

**Sampling bias is confounded with the rainfall gradient.** Collector effort is
not uniform across Australia — wetter regions are systematically better sampled
because they are more accessible and more botanically diverse. Because effort
covaries with rainfall, observed richness conflates true biological signal with
detection probability. To separate these, we use total vascular plant records
per cell as a direct measure of sampling effort, diagnose the effort–richness
relationship via species accumulation curves, and apply a rarefaction-based
correction that resamples the full record pool at standardised effort levels
(100 and 500 total records per cell) using a power-law accumulation model,
projecting expected richness for all four origin × life-form groups
simultaneously. These effort-corrected estimates are used as a sensitivity check
against the raw-count models.

```{r}
#| label: fig-resample
include_graphics(c(tar_read(figS_resample_png_100), tar_read(figS_resample_png_500)))
```

```{r}
#| label: tbl-slopes
#| tbl-cap: "Raw against effort-standardised slopes."
tar_read(slopes_shift_table) |> kable()
```

## Figures E3 and E4. Richness coloured by population accessibility

```{r}
#| label: fig-e3-e4
include_graphics(c(tar_read(figE3_png), tar_read(figE4_png)))
```

## Difference in annual fraction, native against introduced

Not currently included in the paper.

```{r}
#| label: fig-4b
include_graphics(tar_read(fig4b_png))
```

# Values quoted in the manuscript

Each row gives the value as written in the manuscript, the value the pipeline
currently produces, and whether the two differ.

```{r}
#| label: tbl-reported
tar_read(reported_values_table) |>
  mutate(across(where(is.numeric), \(x) signif(x, 5))) |>
  kable()
```
