---
title: "Dataset construction"
format: html
execute:
  echo: false
  warning: false
---

```{r}
#| label: setup
#| include: false
library(targets)
library(dplyr)
library(knitr)
```

This document describes how the analysis dataset is assembled and reports the
result of each stage. It performs no computation — the pipeline in `_targets.R`
does the work, and every number below is read from it with `tar_read()`.

Run `targets::tar_make()` to build the dataset. `targets::tar_visnetwork()`
shows the dependency graph.

```{r}
#| label: check
if (length(tar_outdated()) > 0) {
  warning("Some targets are out of date. Run targets::tar_make() first.")
}
```

## Occurrence records

Occurrences come from a single frozen GBIF download
([0001494-250914085247600](https://www.gbif.org/occurrence/download/0001494-250914085247600)),
so the input resolves to the same 24,385,439 records indefinitely rather than to
whatever GBIF currently holds.

Records are filtered to Australian, present, non-fossil, non-cultivated records
with adequate coordinate precision (better than 0.05°) and uncertainty (under
10 km), collected from 1900 onward, at species rank or below. Two institutions
are excluded for known georeferencing problems. Records at country and state
centroids, capital cities and biodiversity institutions are then removed, using
buffers of 5 km, 2 km and 2 km respectively, and the result is deduplicated by
location, species and year.

```{r}
#| label: tbl-records
tibble::tibble(
  stage = c("Raw download", "After cleaning and deduplication",
            "After joining traits, taxonomy and climate"),
  records = c(24385439L,
              nrow(tar_read(gbif_clean)),
              nrow(tar_read(data_analysis_individuals)))
) |> mutate(records = format(records, big.mark = ",")) |> kable()
```

## Taxonomy and origin

Names are aligned to the Australian Plant Census using APCalign, against a
pinned release (set in `R/align_taxonomy.R`). Alignment assigns each name an
accepted name and a native or introduced status; taxa that are not APC-accepted
are dropped.

Pinning the release matters: changing it moves species between native and
introduced and shifts richness counts, so it is a scientific decision rather
than a maintenance one.

## Life history

Life-history classifications come from AusTraits v7.0.0 (dataset `Wenk_2023`),
re-aligned to the same APC release, since the AusTraits names were aligned
against an earlier one. Taxa recorded as both annual and perennial —
facultative annuals — are counted as annual.

```{r}
#| label: tbl-lifehistory
tar_read(species_life_history) |> count(life_history) |> kable()
```

## Climate and population covariates

WorldClim 2.1 at 2.5 arcmin supplies mean annual temperature (bio1), mean annual
precipitation (bio12), precipitation of the driest month (bio14), precipitation
seasonality (bio15) and precipitation of the driest quarter (bio17).

Population density comes from the ABS Australian Population Grid 2011, summed
onto the analysis grid and converted to people per km². Population accessibility
is a Gaussian kernel-weighted sum of the square root of density, computed at
100, 250 and 500 km bandwidths; the 250 km band is used in the analysis.

## Analysis grid

A regular 0.75° grid in EPSG:4326 covering Australia. Climate values are
extracted at cell centres by bilinear interpolation — the grid aligns exactly
with the WorldClim layers, so centres fall on the corner between four climate
cells, and interpolation averages those four.

```{r}
#| label: tbl-grid
grid <- tar_read(grid_square)
tibble::tibble(
  property = c("Resolution", "Columns x rows", "Cells", "Cells with climate and population data"),
  value = c("0.75 degrees",
            paste(terra::ncol(grid), "x", terra::nrow(grid)),
            as.character(terra::ncell(grid)),
            as.character(nrow(tar_read(grid_climate))))
) |> kable()
```

## Assembled dataset

Each retained occurrence carries its species, origin, life history, climate,
population value and grid cell. Richness is then tabulated per cell for each of
the four origin × life-history groups.

```{r}
#| label: tbl-groups
tar_read(data_analysis_individuals) |>
  filter(!is.na(life_history), establishment_means %in% c("native", "introduced")) |>
  distinct(species, establishment_means, life_history) |>
  count(establishment_means, life_history, name = "species") |>
  kable()
```

## Global climate reference

A 1° global climate grid is also built, used only to place Australian climate
space in a global context (Figure S1).
